Reproducible Research: Peer Assessment 2
==========================================
Created by Robert Tuck June 18, 2015

## Impact of Severe Weather Events on Public Health and Economy in the United States

### Synonpsis  
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

From the analysis of the data, it appears that Tornado events are the most harmful with respect to population health and Flood events have the greatest economic impact.


### Basic settings
```{r}
library(R.utils)
library(ggplot2)
require(plyr)
library(car)
library(dplyr)
```

### Data Processing
```{r}
# download file from URL
if (!file.exists("storm.csv.bz2")) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
        "storm.csv.bz2", method = "curl")
}
# unzip file
if (!file.exists("storm.csv")) {
    bunzip2("storm.csv.bz2", "storm.csv", remove = FALSE)
}
# load data into R
storm <- read.csv("storm.csv")
head(storm)
```
Get a subset of the data with columns we are interested in.
```{r}
### subset the data into the columns of interest
datacols <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
storm_data <- storm[datacols]
```

Convert the PROPDMGEXP codes into numeric values.  See https://rpubs.com/flyingdisc/PROPDMGEXP for an analysis about converting these values.  Then do a smilar conversion for CROPDMGEXP.


```{r}
# unique(storm_data$PROPDMGEXP)
# Function to recode a data column from one set a values to another
# See https://susanejohnston.wordpress.com/2012/10/01/find-and-replace-in-r-part-2-how-to-recode-many-values-simultaneously/ for additional info

recoderFunc <- function(data, oldvalue, newvalue) {
  # convert any factors to characters
  if (is.factor(data))     data     <- as.character(data)
  if (is.factor(oldvalue)) oldvalue <- as.character(oldvalue)
  if (is.factor(newvalue)) newvalue <- as.character(newvalue)
  
  # create the return vector
  newvec <- data
  
  # put recoded values into the correct position in the return vector
  for (i in unique(oldvalue)) newvec[data == i] <- newvalue[oldvalue == i]
  
  newvec
  }

prop_exp_from <- c("H","h","K","k","M","m","B","b","+","-","?","0","1","2","3","4","5","6","7","8","")
prop_exp_to <- c(100, 100, 1000, 1000, 1000000, 1000000, 1000000000, 100000000, 1, 0, 0, 1, 10, 100, 1000, 10000, 100000, 100000, 10000000, 100000000, 0)

storm_data$PROPEXP <- as.numeric(recoderFunc(storm_data$PROPDMGEXP, prop_exp_from, prop_exp_to))

crop_exp_from <- c("?", "K", "k", "m", "M", "B", "0", "2", "")
crop_exp_to <- c(0, 1000, 1000, 1000000, 1000000, 1000000000, 1, 100, 0)
storm_data$CROPEXP <- as.numeric(recoderFunc(storm_data$CROPDMGEXP, crop_exp_from, crop_exp_to))
```

Combine the values from PROPDMG and PROPEXP into PROPDMGVALUE by multiplying the PROPDMG by the PROPEXP to get the actual value of the proerty damage.
```{r}
storm_data$PROPDMGVALUE <- storm_data$PROPDMG * storm_data$PROPEXP
```
Combine the values from CROPDMG and CROPEXP into CROPDMGVALUE by multiplying the CROPDMG by the CROPEXP to get the actual value of the crop damage.
```{r}
storm_data$CROPDMGVALUE <- storm_data$CROPDMG * storm_data$CROPEXP
```
Add the values of the property damage (PROPDMGVALUE) and crop damage (CROPDMGVALUE) and store in a new column, TOTALDMGVALUE.
```{r}
storm_data$TOTALDMGVALUE <- storm_data$PROPDMGVALUE + storm_data$CROPDMGVALUE
```

Get total values of fatalities, injuries and econmic costs by aggregating values based on the type of event (EVTYPE)
```{r}
fatalities <- aggregate(FATALITIES ~ EVTYPE, data=storm_data, sum)
injuries <- aggregate(INJURIES ~ EVTYPE, data=storm_data, sum)
propdmg <- aggregate(PROPDMGVALUE ~ EVTYPE, data=storm_data, sum)
cropdmg <- aggregate(CROPDMGVALUE ~ EVTYPE, data=storm_data, sum)
totaldmg <- aggregate(TOTALDMGVALUE ~ EVTYPE, data=storm_data, sum)
# create a new column in the totaldmg to get costs in billions of dollars
totaldmg$VALUE_BILLION <- totaldmg$TOTALDMGVALUE / 1000000000
```

### Results

#### Across the United States, Which types of events are nost harmful with respect to population health?

```{r}
top10_fatalities <- head(fatalities[order(-fatalities$FATALITIES),],10)
top10_fatalities

top10_injuries <- head(injuries[order(-injuries$INJURIES),], 10)
top10_injuries

ggplot(data=top10_fatalities, aes(x=reorder(EVTYPE,-FATALITIES), y=FATALITIES)) +
     geom_bar(stat="identity") + 
     xlab("Event Type") + 
     ylab("Number of Fatalities") + 
     ggtitle("Top 10 Weather Events with Highest Fatalities 1950 - 2011") +
     theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data=top10_injuries, aes(x=reorder(EVTYPE,-INJURIES), y=INJURIES)) +
     geom_bar(stat="identity") + 
     xlab("Event Type") + 
     ylab("Number of Injuries") + 
     ggtitle("Top 10 Weather Events with Highest Injuries 1950 - 2011") +
     theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Across the United States, which types of events have the greatest economic consequences?

```{r}
top10_totaldmg <- head(totaldmg[order(-totaldmg$TOTALDMGVALUE),],10)
top10_totaldmg

ggplot(data=top10_totaldmg, aes(x=reorder(EVTYPE,-VALUE_BILLION), y=VALUE_BILLION)) +
     geom_bar(stat="identity") + 
     xlab("Event Type") + 
     ylab("Cost (Billions of US Dollars)") + 
     ggtitle("Top 10 Weather Events with Highest Economic Cost 1950 - 2011") +
     theme(axis.text.x = element_text(angle = 45, hjust = 1))
```